---
title: "Intro to Data Viz"
format: html
editor_options: 
  chunk_output_type: console
---

## Set up
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(janitor) # expedite cleaning and exploring data
library(scales) # scale functions for visualization
library(leaflet) # interactive maps
library(DT) # interactive tables
```

## Load working file - Read data
```{r}
escape_raw = read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")
```

The data is daily salmon escapement counts in Alaska 1921-2017 from OceanK database

## Exploring data
```{r}
# Doing this means it runs but it does not show on render
#| eval: false

## Check out column names
colnames(escape_raw)

## Peak at each column and class
glimpse(escape_raw)

## From when to when
range(escape_raw$sampleDate)

## Which species?
unique(escape_raw$Species)
```


## Cleaning data
```{r}
escape = escape_raw %>% 
  clean_names()

colnames(escape)
```


Calculate the Annual escapements by species and sasap regions
```{r}
anual_esc = escape %>% 
  separate(sample_date, c("year", "month", "day"), sep = "-") %>% 
  mutate(year = as.numeric(year)) %>%
  group_by(species, sasap_region, year) %>% 
  summarize(escapement = sum(daily_count)) %>% 
  filter(species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink")) %>% 
  ungroup()

unique(anual_esc$species)

head(anual_esc)
```

## Plotting data
```{r}
ggplot(data = anual_esc,
       mapping = aes(x = species, y = escapement)) +
  geom_col()


```

Making bars blue
```{r}
ggplot(anual_esc,
       aes(x = species, y = escapement,)) +
  geom_col(fill = "blue")
  
```

Adding regions to our plot
```{r}
ggplot(anual_esc,
       aes(x = species, y= escapement, fill = sasap_region)) + 
  geom_col()

```


## Create multiple plots
```{r}
## Subset with data from years 2000 to 2016

anual_esc_2000s <- anual_esc %>%
    filter(year %in% c(2000:2016))

## Quick check
unique(anual_esc_2000s$year)
```

Plotting
```{r}
ggplot(anual_esc_2000s,
       aes(x = year, y = escapement, color = species)) + 
  geom_line() + 
  geom_point() +
  facet_wrap(~ sasap_region, scales = "free_y")

```

Customizing plot
```{r}
ggplot(anual_esc_2000s,
       aes(x = year,
           y = escapement,
           color = species)) +
    geom_line() +
    geom_point() +
    facet_wrap( ~ sasap_region,
                scales = "free_y") +
    labs(title = "Anual Salmon Escapement by Region",
         y = "Escapement") +
    theme_bw() + 
    theme(legend.position = "bottom", 
          legend.title = element_blank(),
          axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))
    scale_y_continuous(labels = comma)
```

Saving plot
```{r}
ggsave("plots/anual_salmon_esc_region.jpg", width = 8, height = 6, units = "in")
```


## Interactive viz
Subset data
```{r}
location <- escape %>% 
  distinct(location, latitude, longitude) %>% 
drop_na()
```

```{r}
datatable(location)
```


Maps with leaflet (OSM)
```{r}
leaflet(location) %>%
    addTiles() %>%
    addMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ location
    )
```

Maps with leaflet (WMS)
```{r}
leaflet(location) %>%
    addWMSTiles(
        "https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?request=getmap&service=wms&BBOX=-90,-180,90,360&crs=EPSG:4326&format=image/jpeg&layers=gebco_latest&width=1200&height=600&version=1.3.0",
        layers = 'GEBCO_LATEST',
        attribution = "Imagery reproduced from the GEBCO_2022 Grid, WMS 1.3.0 GetMap, www.gebco.net"
    ) %>%
    addCircleMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ location,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1
    )
```

