---
title: "Mini Literate Analysis"
format: html
editor_options: 
  chunk_output_type: console
---

## Introduction

### About the data

This data set contains North Pole Environmental Observatory Bottle Chemistry data. This data was downloaded from the [Arctic Data Center](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X)

### Setup

```{r}
library(readr)
```

### Read Reada

```{r}
bg_chem <- read_csv("data/raw/data/BGchem2008data.csv")
```

Exploring data

```{r}
colnames(bg_chem)
summary(bg_chem)
unique(bg_chem$Date)
```

## Analysis

### Calculate Summary Statistics

```{r}
nitate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)

```

### Calculate the mean Readfield ratio

```{r}
ratio <- (nitate + nitrite + amm)/phos
```

### Plot Ratio

<!-- This is a comment that will not appear in render -->

```{r}
plot(bg_chem$P, bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4) #this is a sketch plot
```

## Conclusion

The Redfield ratio for this dataset is approximately: `r round(ratio)`
